---
title: "Getting Started with CatchAll"
author: "Kendrick Li"
output: pdf_document
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEncoding[utf8]{inputenc}
---

```{r setup, include=FALSE, cache=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```


This package contains functions that estimate the species richness with finite exponential-mixed Poisson models (up to the three-component exponential mixed-Poisson model) and other state-of-art species richness estimators.
## Installation ##


```{r}
library(devtools)
install_github("KenLi93/CatchAll")

```

## Basic Usage ##


```{r}
library(CatchAll)
library(breakaway)
data(apples)
```

Poisson Model:
```{r}
Poisson_model(apples, cutoff = 20)
```

Geometric model (or single exponential-mixed Poisson model):
```{r}
geometric_model(apples, cutoff = 20)
```

Two- or three-component exponential mixture mixed Poisson model:
```{r}
two_geometric_model(apples, cutoff = 50)
```

```{r}
three_geometric_model(apples, cutoff = 70)
```

An integrated function for simultaneously viewing the results of all 
above models for different cut-off of species frequency counts.

```{r} 
apples_results <- all_parametric_model(apples)
head(apples_results, 10)
```

The richness estimates in CatchAll are displayed as follows:

```{r}
apples_best_models <- catch_all(apples)
apples_best_models
```

